---
title: "Weakly Informative Examples"
output:
  html_document:
    df_print: paged
---

# Do beautiful parents have more daughters?

```{r message=FALSE, warning=FALSE}
library(rstan)

data_list <- list(
  n_attractive = 300,
  n_daughters_attractive = 170,
  n_nonattractive = 2700,
  n_daughters_nonattractive = 1310
)

dat_frm <- matrix(c(170,300,0.567, 1310,2700, 0.485), 2, 3, byrow = T)
rownames(dat_frm) <- c("Beautiful Parents", "Other Parents")
colnames(dat_frm) <- c("Daughters", "Total", "Proportion")

print(dat_frm)
```

## Model with uniform prior

$y_1\sim Binomial(300,p_1)$,$y_0 \sim Binomial(2700, p_0)$, 
```{r}
stan_code_uniform<- "
data {
  int<lower=0> n_attractive;      // Total number of attractive parents
  int<lower=0> n_daughters_attractive;  // Number of attractive parents with daughters
  
  int<lower=0> n_nonattractive;      // Total number of non-attractive parents
  int<lower=0> n_daughters_nonattractive;  // Number of non-attractive parents with daughters
}

parameters {
  real<lower=0, upper=1> p_nonattractive;  // Proportion of non-attractive parents having daughters
  real<lower=0, upper=1> p_attractive;      // 
}

transformed parameters {
  real<lower=-1, upper=1> difference =p_attractive- p_nonattractive;  // Calculate p_attractive from difference
}

model {
  // Uniform priors
  p_nonattractive ~ uniform(0, 1);
  p_attractive ~ uniform(0, 1);

  // Binomial likelihood
  n_daughters_attractive ~ binomial(n_attractive, p_attractive);
  n_daughters_nonattractive ~ binomial(n_nonattractive, p_nonattractive);
}


"

model2 <- stan_model(model_code=stan_code_uniform)
# Sample from the model
fit2 <- sampling(model2, data=data_list, chains=4, iter=20000, warmup=1000, cores=4, refresh=0)
print(fit2, digits=3)
```
% of MCMC with diff>0

```{r}
vec = extract(fit2)$difference
pct = sum(vec>0)/length(vec)
print(pct)
```

Plot histogram for posterior MCMC sample for difference

```{r}

print(hist(extract(fit2)$difference, breaks = 50, main = 'Model 1, MCMC sample for diff', xlab='diff'))
```




## Model with weakly informative prior on difference

$y_1\sim Binomial(300,p_1)$,$y_0 \sim Binomial(2700, p_0)$, reparameterized to $p_0$ and $p_{diff} = p_1-p_0$

Assuming the difference p1-p0 is less than 1.5% with 95% probability. The intuition is that p0 is 0.485,and we don't believe p1 is greater than 0.5

Let's plot the prior distribution of p_1:


```{r}
# Prior of p_att (aka P_1)
p_1_prior = function(x) {
  if (0 < x & x < 1) {
    return(pnorm((1-x)/sqrt(0.007))-pnorm(-x/sqrt(0.007)))
  } else {
    return(0)
  }
}

x = seq(-1,2,0.01)
y=sapply(x, p_1_prior)
plot(x,y, type='l', main = 'Prior Distribution of p_att', ylab="Density", xlab="p_att")#, xaxs='i', yaxs='i')

# Prior of p_oth (aka P_0)
p_0_prior = function(x) {if (0<x & x<1) {return(1)} else {return(0)}}
x = seq(-1,2,0.01)
y=sapply(x, p_0_prior)
plot(x,y, type='l', main = 'Prior Distribution of p_oth', ylab="Density", xlab="p_oth")#, xaxs='i', yaxs='i')

```





```{r}
library(rgl)
library(colorRamps)

joint_prior = function(x) {
  if(0<x[1] & x[1]<1) {
    return(dnorm(x[2], x[1], 0.007))
  } else {
    return(0)
  }
}
df1 = as.data.frame(expand.grid(seq(0,1,0.01), seq(0,1,0.01)))
names(df1) = c('p_oth', 'p_att')
df1$joint_prior = apply(df1, 1, joint_prior)

rbPal <- colorRampPalette(c('grey', 'red'))
df1$col <- rbPal(10)[as.numeric(cut(df1$joint_prior,breaks = 10))]

plot3d( 
  x=df1$`p_oth`, y=df1$`p_att`, z=df1$`joint_prior`, 
  type = 'l', 
  col = df1$col,
  radius = .1,
  xlab="p_oth", ylab="p_att", zlab="Density")

```

What % of the density of p_att aka p_1 is less than 0 and greater than 1?

```{r}
run_draw = function(x) {
  p_oth = runif(1)
  p_att = rnorm(1, p_oth, sqrt(0.007))
  return(p_att)
}

prior_draws = sapply(1:100000, run_draw)
print("Proportion of p_att distirbution <0:")
sum(prior_draws<0)/length(prior_draws)
print("Proportion of p_att distirbution >1:")
sum(prior_draws>1)/length(prior_draws)

```

Run the model with weakly informative priors:


```{r}
stan_code_weakly_inform <- "
data {
  int<lower=0> n_attractive;      // Total number of attractive parents
  int<lower=0> n_daughters_attractive;  // Number of attractive parents with daughters
  
  int<lower=0> n_nonattractive;      // Total number of non-attractive parents
  int<lower=0> n_daughters_nonattractive;  // Number of non-attractive parents with daughters
}

parameters {
  real<lower=0, upper=1> p_nonattractive;  // Proportion of non-attractive parents having daughters
  real<lower=-1, upper=1> difference;      // p_attractive - p_nonattractive
}

transformed parameters {
  real<lower=0, upper=1> p_attractive = p_nonattractive + difference;  // Calculate p_attractive from difference
}

model {
  // Uniform priors
  p_nonattractive ~ uniform(0, 1);
  difference ~ normal(0, 0.007);
  // Binomial likelihood
  n_daughters_attractive ~ binomial(n_attractive, p_attractive);
  n_daughters_nonattractive ~ binomial(n_nonattractive, p_nonattractive);
}


"

model4 <- stan_model(model_code=stan_code_weakly_inform)


# Sample from the model
fit4 <- sampling(model4, data=data_list, chains=4, iter=20000, warmup=1000, cores=4, refresh=0)
print(fit4, digits=3)
```


% of MCMC with diff>0

```{r}
vec = extract(fit4)$difference
pct = sum(vec>0)/length(vec)
print(pct)
```

Plot histogram of posterior dist for difference:

```{r}
print(hist(extract(fit4)$difference, breaks = 50, main = 'Model 2, MCMC sample for diff', xlab='diff'))

```







```{r}
stan_code_weakly_inform <- "
data {
  int<lower=0> n_attractive;      // Total number of attractive parents
  int<lower=0> n_daughters_attractive;  // Number of attractive parents with daughters
  
  int<lower=0> n_nonattractive;      // Total number of non-attractive parents
  int<lower=0> n_daughters_nonattractive;  // Number of non-attractive parents with daughters
}

parameters {
  real<lower=0, upper=1> p_nonattractive;  // Proportion of non-attractive parents having daughters
  real<lower=-1, upper=1> difference;      // p_attractive - p_nonattractive
}

transformed parameters {
  real<lower=0, upper=1> p_attractive = p_nonattractive + difference;  // Calculate p_attractive from difference
}

model {
  // Uniform priors
  p_nonattractive ~ uniform(0, 1);
  difference ~ normal(0, 0.007);
  // Binomial likelihood
  n_daughters_attractive ~ binomial(n_attractive, p_attractive);
  n_daughters_nonattractive ~ binomial(n_nonattractive, p_nonattractive);
}


"

model4 <- stan_model(model_code=stan_code_weakly_inform)


# Sample from the model
fit4 <- sampling(model4, data=data_list, chains=4, iter=2000, warmup=1000, cores=4, refresh=0)
print(fit4, digits=3)
```

fit4



```{r}
# Run this only once! This accrues the results from all simulations.
#simulation_results = list()
```


```{r}

simulation_run = function(prior_diff_var = 0.07, true_p_attractive = 0.49, true_p_nonattractive = 0.49, nsim = 2, 
                          simulation_results_list = simulation_results) {
    
  stan_code_weakly_inform <- paste0("
  data {
    int<lower=0> n_attractive;      // Total number of attractive parents
    int<lower=0> n_daughters_attractive;  // Number of attractive parents with daughters
    
    int<lower=0> n_nonattractive;      // Total number of non-attractive parents
    int<lower=0> n_daughters_nonattractive;  // Number of non-attractive parents with daughters
  }
  
  parameters {
    real<lower=0, upper=1> p_nonattractive;  // Proportion of non-attractive parents having daughters
    real<lower=-1, upper=1> difference;      // p_attractive - p_nonattractive
  }
  
  transformed parameters {
    real<lower=0, upper=1> p_attractive = p_nonattractive + difference;  // Calculate p_attractive from difference
  }
  
  model {
    // Uniform priors
    p_nonattractive ~ uniform(0, 1);
    difference ~ normal(0, ", prior_diff_var, ");
    // Binomial likelihood
    n_daughters_attractive ~ binomial(n_attractive, p_attractive);
    n_daughters_nonattractive ~ binomial(n_nonattractive, p_nonattractive);
  }
  ")
  start_time <- Sys.time()
  model5 <- stan_model(model_code=stan_code_weakly_inform)
  end_time <- Sys.time()
  print(end_time-start_time)
  
  sim_vector = NULL
  for (i in 1:nsim) {
    print(i) # print every 100 to show progress
    start_time <- Sys.time()
      data_list <- list(
        n_attractive = 300, n_daughters_attractive = rbinom(1, 300, true_p_attractive),
        n_nonattractive = 2700, n_daughters_nonattractive = rbinom(1, 2700, true_p_nonattractive)
      )
    
    # Sample from the model
    fit5 <- sampling(model5, data=data_list, chains=4, iter=3000, warmup=1000, cores=4, refresh=0)  
    # record 1 of the CI excludes 0 on the lower side, and 0 otherwise
    lower_ci_bound = summary(fit5)$summary["difference","2.5%"]
    sim_vector = c(sim_vector, 0<lower_ci_bound)
    end_time <- Sys.time()
    print(end_time-start_time)
  }
  
  simID = paste0('prior_diff_var = ', prior_diff_var,', true_p_attractive = ', true_p_attractive, ', true_p_nonattractive = ', true_p_nonattractive)
  
  simulation_results_list[[simID]] = c(simulation_results_list[[simID]], sim_vector)
  return(simulation_results_list)
}


```


```{r}

# Type 1 error
simulation_results = simulation_run(prior_diff_var = 0.007, true_p_attractive = 0.59, true_p_nonattractive = 0.49, nsim = 10000)

```


```{r}

# Type 1 error
simulation_results = simulation_run(prior_diff_var = 100, true_p_attractive = 0.49, true_p_nonattractive = 0.49, nsim = 10000)
simulation_results = simulation_run(prior_diff_var = 0.007, true_p_attractive = 0.49, true_p_nonattractive = 0.49, nsim = 10000)

# Power for true_p_attractive = 0.59, true_p_nonattractive = 0.49
simulation_results = simulation_run(prior_diff_var = 100, true_p_attractive = 0.59, true_p_nonattractive = 0.49, nsim = 10000)
simulation_results = simulation_run(prior_diff_var = 0.007, true_p_attractive = 0.59, true_p_nonattractive = 0.49, nsim = 10000)

```

