---
title: "Election Prediction Model with Prior Setup"
output: html_document
date: "2024-03-29"
---


# Goal

Estimate national and state-level trends in support for the candidates

# How

Partially pool two separate predictions:

- a forecast based on historically relevant economic and political factors. (juneapp + q2gdp)
- information from state and national polls during the election season

# Data

https://github.com/TheEconomist/us-potus-model/blob/master/data/

# Model

For individual poll $i$,

$$
\begin{equation}
y_i \sim Binomial(\theta_i, n_i)
\end{equation}
$$

## Start with a simple model

If poll $i$ is a state-level poll,

$$
\begin{equation}
log(\frac{\theta_i}{1-\theta_i}) = \mu^{b}_{s[i],t[i]}
\end{equation}
$$

- $\mu^{b}_{s,t}$, which represents the underlying support for the Democrat in state $s$ at time $t$

If poll $i$ is a national-level poll,

$$
\begin{equation}
log(\frac{\theta_i}{1-\theta_i}) = \sum_{s=1}^S w_s\mu^{b}_{s,t[i]}
\end{equation}
$$

prior:

$$
\begin{equation}
\begin{aligned}
\mu_{\mathrm{T}}^{\mathrm{b}} & \sim \operatorname{MVN}\left(\mu_{t.f.c.}, b_T*\Sigma\right)\\
\mu_{\mathrm{t}}^{\mathrm{b}} \mid \mu_{\mathrm{t}-1}^{\mathrm{b}} &\sim \operatorname{MVN}\left(\mu_{\mathrm{t}-1}^{\mathrm{b}}, b_0*\Sigma\right)
\end{aligned}
\end{equation}
$$

-  $\mu_{t.f.c.}$: prediction from time-for-change model

### How to select $b_T, b_0, \Sigma$

Use Informative Prior

Construct $\Sigma$ 

1. Estimate the correlation matrix using demographics data from different states.

2. Shrink the pairwise correlation by 10\% to reflect uncertainty 

3. Assume equal standard deviation 7\% for each state, the national s.d. is about 5 \%.

Construct $b_T$ 

1. At the beginning of the poll, let $b_T=1$; at the election day, use $b_T=0.36$ (national s.d. becomes 60\% of  the original one , that is, 5*0.6= 3 \%)



```{r}
fit_rmse_day_x <- function(x){0.03 +  (10^-6.6)*(x)^2} # fit to error from external script
plot(0:300, fit_rmse_day_x(0:300)/0.05,xlab="time to election",ylab="scale",type="l")
```

2. Transform to the logit scale. $b_T = b_T * 4$

```{r}
a <- seq(0,1,length.out=1000)
plot(a,log(a/(1-a)),xlab="x",ylab="y=logit(x)",type="l", main="y ~= 4x-2")
abline(-2,4,col=2)
```


Construct $b_0$

1. Use 1/sqrt(300), roughly 0.05. (5*0.05=0.25\%). The change will be within 0.5\% between neighboring days.

2. Transform to the logit scale. $b_0 = b_0 * 4$



## Consider bias $\alpha_i$ from different sources

state-level poll:

$$
\begin{equation}
log(\frac{\theta_i}{1-\theta_i}) = \mu^{b}_{s[i],t[i]} + \alpha_i
\end{equation}
$$
national-level poll:

$$
\begin{equation}
log(\frac{\theta_i}{1-\theta_i}) = \sum_{s=1}^S w_s\mu^{b}_{s,t[i]}+ \alpha_i
\end{equation}
$$

bias decomposition:

$$
\begin{equation}
\alpha_{\mathrm{i}}=\mu_{\mathrm{p}[\mathrm{i}]}^{\mathrm{c}}+\mu_{\mathrm{r}[\mathrm{i}]}^{\mathrm{r}}+\mu_{\mathrm{m}[\mathrm{i}]}^{\mathrm{m}}+\mathrm{z_i} \epsilon_{\mathrm{t}[\mathrm{i}]}
\end{equation}
$$

- $\mu^c$: house effects. A polling organization's results lean in favor of one party.
- $\mu^r$: polling population effects. Effects caused by different population. (e.g. likely voters and registered voters)
- $\mu^m$: polling mode effects. (e.g. live caller or online)
- $z_i$: equal to 1 if the pollster does not adjust for partisanship and 0 otherwise
- $\epsilon_t$: non-response bias at time $t$

prior:

$$
\begin{equation}
\begin{aligned}
\mu^{\mathrm{c}}_c & \sim \operatorname{Normal}\left(0, \sigma^{\mathrm{c}}\right) \\
\mu^{\mathrm{r}}_r & \sim \operatorname{Normal}\left(0, \sigma^{\mathrm{r}}\right) \\
\mu^{\mathrm{m}}_m & \sim \operatorname{Normal}\left(0, \sigma^{\mathrm{m}}\right)
\end{aligned}
\end{equation}
$$
$$
\begin{equation}
\begin{gathered}
\epsilon_1 \mid \rho, \sigma^{\epsilon} \sim \operatorname{Normal}\left(0, \frac{1}{\sqrt{1-\rho^2}} \sigma^{\epsilon}\right) \\
\epsilon_{\mathrm{t}} \mid \epsilon_{\mathrm{t}-1}, \rho, \sigma^{\epsilon} \sim \operatorname{Normal}\left(\rho \epsilon_{\mathrm{t}-1}, \sigma^{\epsilon}\right) \text { for } \mathrm{t}=2, \ldots, \mathrm{T} \\
\rho \sim \operatorname{Nnormal}(0.7,0.1) .
\end{gathered}
\end{equation}
$$
This prior put most of the probability mass on stationary series, as the non-response bias is not very likely to follow a random walk.

prior (logit scale):

- $\sigma_c = 0.06$. 1.5\% s.d. for house effects

- $\sigma_r = 0.04$. 1\% s.d. for population effects

- $\sigma_m = 0.04$. 1\% s.d. for population effects

- $\sigma_e = 0.02$. 0.5\% s.d. for non-response bias.

## Consider measurement error $\zeta_i$


state-level poll:

$$
\begin{equation}
log(\frac{\theta_i}{1-\theta_i}) = \mu^{b}_{s[i],t[i]} + \alpha_i + \zeta^{state}_i
\end{equation}
$$

national-level poll:

$$
\begin{equation}
log(\frac{\theta_i}{1-\theta_i}) = \sum_{s=1}^S w_s\mu^{b}_{s,t[i]}+ \alpha_i+\zeta^{national}_i
\end{equation}
$$

prior: 

$$
\begin{equation}
\begin{aligned}
\zeta_{\mathrm{i}}^{\text {national }} & \sim \operatorname{Normal}\left(0, \sigma^{\text {national }}\right) \\
\zeta_{\mathrm{i}}^{\text {state }} & \sim \operatorname{Normal}\left(0, \sigma^{\text {state }}\right) 
\end{aligned}
\end{equation}
$$

- $\sigma^{national} = 0.04$. 1\% s.d. for national measuring error

- $\sigma^{national} = 0.04$. 1\% s.d. for state measuring error

## Consider polling error


state-level poll:

$$
\begin{equation}
log(\frac{\theta_i}{1-\theta_i}) = \mu^{b}_{s[i],t[i]} + \alpha_i + \zeta^{state}_i + \xi_{s[i]}
\end{equation}
$$

national-level poll:

$$
\begin{equation}
log(\frac{\theta_i}{1-\theta_i}) = \sum_{s=1}^S w_s\mu^{b}_{s,t[i]}+ \alpha_i+\zeta^{national}_i + 
\sum_{s=1}^S w_s \xi_s
\end{equation}
$$


prior:

$$
\begin{equation}
\xi \sim \operatorname{MVN}\left(0, b_{\xi}*\Sigma\right)
\end{equation}
$$


Construct $b_{\xi}$

1. $b_{\xi}$ is selected to be 0.25 because this will make the prior s.d. for national polling error $\sqrt{w^{T}\Sigma w}$ roughly 1.3\% (prior information).

```{r}
err <- c(0.5, 1.9, 0.8, 7.2, 1.0, 1.4, 0.1, 3.3, 3.4, 0.9, 0.3, 2.7, 1.0) / 2 
sqrt(mean(err^2))
```


2. Transform to the logit scale. $b_{\xi} = b_{\xi} * 4$